Optimizing Structural Properties Of Neural Networks With Genetic Algorithms

### 2015-08-06 paper/paper.tex
commented some stuff out

Since calculating the fitness of a genotype requires the evaluation of many
different phenotypes and because of the resulting stochastic nature of the
problems we faced, we decided to use genetic algorithms for the
optimization process.

@@

\paragraph{Tasks}
For the optimization to work properly, the tasks presented to the networks
must be chosen reasonably; they must be learnable by FORCE (at least in
principle) and must be compatible with technical aspects like the
integration time step. 
While it was shown in \citep{FORCE} that large recurrent neural networks are
capable of learning very complex tasks, the optimization requires an
extensive number of learning processes, since every single evaluation of
the fitness $F_C$ needs dozens of tested phenotypes (see below). Therefore,
the networks had to be restricted to being quite small ($\approx 100$
neurons) and are consequently less powerful.

@@

Since we expect/hope that the optimized network extends this range of
possible frequencies, we choose the frequencies for the optimization tasks
from the slightly larger interval $[0.6e-2, 4]$ such that their logarithms
are uniformly spaced. This allows for good resolution for both high and low
frequencies.

@@

A measure that is time-shift independent and maximized for functions of
identical shape is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(r, r') = \max_t \frac{\left<r, r'(\cdot
    - t)\right>}{\sqrt{\left< r, r\right> \left<r', r' \right>}}~.
\end{align}
Algorithmically, the signals $r$ and $r'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
This measure becomes problematic for functions that are very close to zero
for a longer time (order of chunk size), but since we are going to work
with superpositions of sinusoidals, this drawback
is unproblematic.

@@

Therefore, we optimized $F_C(G)$ using a genetic algorithm, which is inherently
capable of coping with the stochastic fitness function. In each iteration
of the algorithm, a generation $\mathcal{G}_t = \left\{ G_i~|~1\le i\le N\right\}$
of $N$ individuals (Generators) is converted into a new one, $\mathcal{G}_{t+1}$,
according to the following scheme:
\begin{description}[itemsep=1ex, leftmargin=0pt]
 \item[Evaluation] The fitness $F_i = F_C(G_i)$ of each individual is
     measured using at least 50 samples (and adaptively more if the
     $(F_i)_{1\le i\le N}$ are spaced closely, to at most 100).
 \item[Selection] All individuals are grouped into disjunct pairs for which
     we check which one has the higher fitness. This is repeated $k=10$ times,
     counting the number of wins for each individual. Based thereon we use the
     $p=50\%$ best individuals $\mathcal{G}\Ix{survive}_t$ to build the next
     generation.  By using pair comparisons instead of simple ordering
     according to the fitness, we allow weaker individuals to reach
     the next generation (albeit with low probability), keeping the gene
     pool more diverse.

\item[Reproduction] We then replace the $(1-p)|G_t|$ elements that did not make
it into the next generation by eith \emph{mutating} or \emph{recombining} the 
survivors and get $\mathcal{G}_{t+1}$. This means we either change
    a single parameter randomly or take two individuals and randomly choose
    for each parameter from which parent to take the value.   
\end{description}

@@

- Problems
    * Many magic / ``empiric'' parameters scattered in the model that were not examined exhaustively
    * Only few and uncomprehensive number of test / experiments conducted due to the
      computational cost.
    * Test with several other topologies did not yield concrete results,
      only rough tendencies.
    * Influence of statistics: Learning success depends stronger on choice
      of challenges than parameters (partially) -> must be very careful how to
      interpret the results / should have used fixed array of challenges instead
      of random sampling

>>>>

% TODO we say the same thing later on again. maybe, the comment about the genetic algorithm
% should be moved out of the parenthesis
%Since calculating the fitness of a genotype requires the evaluation of many
%different phenotypes and because of the resulting stochastic nature of the
%problems we faced, we decided to use genetic algorithms for the
%optimization process.

@@

\paragraph{Tasks}
For the optimization to work properly, the tasks presented to the networks
must be chosen reasonably; they must be learnable by FORCE (at least in
principle) and must be compatible with technical aspects like the
integration time step. 
%While it was shown in \citep{FORCE} that large recurrent neural networks are
%capable of learning very complex tasks, the optimization requires an
%extensive number of learning processes, since every single evaluation of
%the fitness $F_C$ needs dozens of tested phenotypes (see below). Therefore,
%the networks had to be restricted to being quite small ($\approx 100$
%neurons) and are consequently less powerful.

@@

Since we expect/hope that the optimized network extends this range of
possible frequencies, we choose the frequencies for the optimization tasks
from the slightly larger interval $[\num{0.6e-2}, 4]$ such that their logarithms
% TODO this number is no longer right: we have low, high and full
are uniformly spaced. This allows for good resolution for both high and low
frequencies.

@@

A measure that is time-shift independent and maximized for functions of
identical shape is the maximum of the cross correlation 
\begin{align}
    f\ix{cor}(r, r') = \max_t \frac{\left<r, r'(\cdot
    - t)\right>}{\sqrt{\left< r, r\right> \left<r', r' \right>}}~.
\end{align}
Algorithmically, the signals $r$ and $r'$ are split into overlapping chunks
which are correlated separately.  This saves computational power and
prevents minute differences in frequency to accumulate into a huge phase
shift, so the phase only has to remain relatively constant over the
timescale of a single chunk (which is about 10 seconds of simulated time).
%This measure becomes problematic for functions that are very close to zero
%for a longer time (order of chunk size), but since we are going to work
%with superpositions of sinusoidals, this drawback
%is unproblematic.

@@

Therefore, we optimized $F_C(G)$ using a genetic algorithm, which is inherently
capable of coping with the stochastic fitness function. In each iteration
of the algorithm, a generation $\mathcal{G}_t = \left\{ G_i~|~1\le i\le N\right\}$
of $N$ individuals (Generators) is converted into a new one, $\mathcal{G}_{t+1}$,
according to the following scheme:
\begin{description}[itemsep=1ex, leftmargin=0pt]
 \item[Evaluation] The fitness $F_i = F_C(G_i)$ of each individual is
     measured using at least 50 samples (and adaptively more if the
     $(F_i)_{1\le i\le N}$ are spaced closely, to at most 100).
 \item[Selection] We choose $p=50\%$ of the most successful members of $G_t$ as survivors.
With a small chance, a weaker generator can survive, in order to keep the gene pool more diverse.

\item[Reproduction] We then replace the $(1-p)|G_t|$ elements that did not make
it into the next generation by either \emph{mutating} or \emph{recombining} the 
survivors and get $\mathcal{G}_{t+1}$. This means we either change
    a single parameter randomly or take two individuals and randomly choose
    for each parameter from which parent to take the value.   
\end{description}

@@

- Problems
    * Only few and uncomprehensive number of test / experiments conducted due to the
      computational cost.
    * Influence of statistics: Learning success depends stronger on choice
      of challenges than parameters (partially) -> must be very careful how to
      interpret the results / should have used fixed array of challenges instead
      of random sampling
